Causal Discovery on Precarity and Depression

### 2025-01-01 draft_v1/draft_v1.qmd
update_draftv1

After preprocessing, the HELIUS dataset comprises 21,628 samples. In addition to the five precariousness factors, we include PHQ-9 scores — both the total sum score and individual symptom scores — to represent depression. In the subsequent causal discovery analysis, we will examine depression both as an aggregated sum score and through its individual symptom-level representations. Refer to @fig-dist for the overall distributions of the variables used in the analysis.

```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%"}
#| fig-cap: >
#|   Distributions of variables with density overlay.  
#|   
#|   *P.emp* = employment precariousness; *P.hou* = housing precariousness; *P.soc* = social precariousness; *S.fin* = recent financial stressors; *S.rel* = recent relational stressors;
#|   *PHQsum* = PHQ-9 sum score; *anh* = anhedonia; *app* = appetite; *con* = concentration; *dep* = depressed mood; *ene* = energy; *glt* = guilty; *mot* = motor; *sui* = suicidal
#| label: fig-dist

@@

## Analysis
We analyze the causal structure using two approaches: one with the PHQ sum score representing overall depression severity, and another with individual symptom scores. The PHQ sum score simplifies the analysis by reducing dimensionality, which is computationally advantageous and provides a broad, interpretable perspective on depression's relationship with precarity factors. In contrast, individual symptom scores offer a nuanced understanding, capturing the heterogeneous ways symptoms respond to precarity factors. However, analyzing individual symptoms poses methodological challenges due to their non-standard distributions and the increased complexity introduced by the higher dimensionality. By employing both approaches, we balance simplicity and granularity, ensuring robustness in our findings: the PHQ sum score captures overarching trends, while symptom-level analysis reveals detailed dynamics, essential for tailoring specific interventions.

To evaluate the sensitivity of the results to the choice of alpha levels and ensure consistency, we test two significance levels --- $\alpha = 0.01, 0.05$. To further ensure robustness, we employ bootstrapping to generate 100 bootstrap samples For each sample, we estimate causal graphs and retain only the edges and directions that appear above predefined thresholds.

The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms, with each combination repeated for 100 bootstrap samples, resulting in a total of 1,600 resulting graphs. For analyses involving individual symptom variables, we streamline the setup by using thresholds of 0.6 and 0.7, reducing the number of bootstrap samples to 30. 
<!-- Additionally, we fix the skeleton of edges among symptom variables based on the common structure estimated across all three algorithms, thereby reducing the computational time required to estimate skeletons. This fixed structure aligns with commonly observed skeleton structures in the literature, ensuring relevance to existing findings (**cite the comp_model paper**).[^2]  -->

To summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression. Refer to @fig-workflow for the analysis workflow.

@@

knitr::include_graphics("img/simsetup.png")

@@

## Precarity as sum score

>>>>

After preprocessing, the HELIUS dataset comprises 21,628 samples. Along with the five precariousness factors, we also compute an overall precarity score as the combined value of these factors. Additionally, PHQ-9 scores are included to represent depression, both as a total sum score and as individual symptom scores. In the subsequent causal discovery analysis, we examine the relationship between depression and precarity using both aggregated sum scores and their individual-level representations. Refer to @fig-dist for the overall distributions of the variables used in the analysis.

```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%"}
#| fig-cap: >
#|   Distributions of variables with density overlay.  
#|   
#|   *P.emp* = employment precariousness; *P.hou* = housing precariousness; *P.soc* = social precariousness; *S.fin* = recent financial stressors; *S.rel* = recent relational stressors; *Precarity* = overall precarity;
#|   *PHQsum* = PHQ-9 sum score; *anh* = anhedonia; *app* = appetite; *con* = concentration; *dep* = depressed mood; *ene* = energy; *glt* = guilty; *mot* = motor; *sui* = suicidal
#| label: fig-dist

@@

## Analysis
We analyze the causal structure using three approaches: examining the relationship (1) between the five individual precarity factors and the PHQ sum score, which represents overall depression severity; (2) between individual symptoms and precarity factors; and (3) between individual symptoms and overall precarity (the sum of the five precarity factors). Using the PHQ sum score simplifies the analysis by reducing dimensionality, offering computational efficiency and a broad, interpretable perspective on the relationship between depression and precarity factors. In contrast, including individual symptoms provide a more detailed understanding, capturing the diverse ways symptoms respond to different precarity factors. However, this approach introduces methodological challenges, such as non-standard distributions and the complexities of increased dimensionality. Finally, examining individual symptoms in relation to the overall precarity score complements the analysis by revealing how symptoms collectively connect to general precarity. By integrating these three approaches, we balance simplicity and granularity, ensuring robust findings. 
<!-- The PHQ sum score and precarity sum score capture overarching trends between overall depression and precarity, while symptom-level analysis with individual precarity factors reveals detailed dynamics, essential for tailoring specific interventions. -->

To address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping. For each of the 100 bootstrap samples, we estimate causal graphs and retain only the edges and directions that exceed predefined thresholds.
The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms. Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm. For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30 to ease computational demands.
<!-- Additionally, we fix the skeleton of edges among symptom variables based on the common structure estimated across all three algorithms, thereby reducing the computational time required to estimate skeletons. This fixed structure aligns with commonly observed skeleton structures in the literature, ensuring relevance to existing findings (**cite the comp_model paper**).[^2]  -->

To summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.  See @fig-workflow for an overview of the analysis workflow.

@@

knitr::include_graphics("img/simsetup2.png")

@@

## Precarity as sum score
Lastly, we examine the relationships between individual symptoms and overall precarity, represented as the sum of five precariousness factors. This approach completes the analysis by providing a perspective on how individual symptoms relate to overall precarity. It complements the previous analysis, which focused on individual symptoms and individual precarity factors, by potentially capturing distributed relationships across different precarity factors through their aggregation into a single overall score.
As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree. Circles at the endpoints of gray dashed edges denote unresolved directionality. Additionally, navy dashed edges highlight relationships that appear exclusively in graphs generated using Gaussian CI tests.

Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges. FCI tends to estimate more undirected edges (circles), while CCI more frequently assigns directed arrows between symptom-symptom relationships. For a detailed breakdown, see [@sec-propmatrix], which shows the proportion of symbols produced by each algorithm.
When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym. In addition to *glt*, *slp*, *anh*, and *sui*, which are previously shown to be connected to individual precarity factors, *mot*, *app*, and *dep* are now identified as having causal relationships with overall precarity. Furthermore, when using the Gaussian CI test, con is also shown to be causally related to overall precarity. Notably, all these connections are directed towards precarity (o→), indicating that overall precarity is not the cause of symptoms. The exception is the bidirectional edge between *sui* <-> *Precarity*, which suggests a connection via latent confounders.
FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity. With fewer variables (reduced from 14 to 9), this approach appears to have slightly more statistical power, enabling algorithms to estimate relationships with greater confidence compared to the previous analysis involving individual symptoms and individual precarity factors.

### 2025-01-01 draft_v1/draft_v1.tex
update_draftv1

\date{2024-12-28}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

@@

After preprocessing, the HELIUS dataset comprises 21,628 samples. In
addition to the five precariousness factors, we include PHQ-9 scores ---
both the total sum score and individual symptom scores --- to represent
depression. In the subsequent causal discovery analysis, we will examine
depression both as an aggregated sum score and through its individual
symptom-level representations. Refer to Figure~\ref{fig-dist} for the
overall distributions of the variables used in the analysis.

@@

\caption{\label{fig-dist}Distributions of variables with density
overlay. \emph{P.emp} = employment precariousness; \emph{P.hou} =
housing precariousness; \emph{P.soc} = social precariousness;
\emph{S.fin} = recent financial stressors; \emph{S.rel} = recent
relational stressors; \emph{PHQsum} = PHQ-9 sum score; \emph{anh} =
anhedonia; \emph{app} = appetite; \emph{con} = concentration; \emph{dep}
= depressed mood; \emph{ene} = energy; \emph{glt} = guilty; \emph{mot} =
motor; \emph{sui} = suicidal}

@@

We analyze the causal structure using two approaches: one with the PHQ
sum score representing overall depression severity, and another with
individual symptom scores. The PHQ sum score simplifies the analysis by
reducing dimensionality, which is computationally advantageous and
provides a broad, interpretable perspective on depression's relationship
with precarity factors. In contrast, individual symptom scores offer a
nuanced understanding, capturing the heterogeneous ways symptoms respond
to precarity factors. However, analyzing individual symptoms poses
methodological challenges due to their non-standard distributions and
the increased complexity introduced by the higher dimensionality. By
employing both approaches, we balance simplicity and granularity,
ensuring robustness in our findings: the PHQ sum score captures
overarching trends, while symptom-level analysis reveals detailed
dynamics, essential for tailoring specific interventions.

To evaluate the sensitivity of the results to the choice of alpha levels
and ensure consistency, we test two significance levels ---
\(\alpha = 0.01, 0.05\). To further ensure robustness, we employ
bootstrapping to generate 100 bootstrap samples For each sample, we
estimate causal graphs and retain only the edges and directions that
appear above predefined thresholds.

The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds
× 2 CI tests), applied across three algorithms, with each combination
repeated for 100 bootstrap samples, resulting in a total of 1,600
resulting graphs. For analyses involving individual symptom variables,
we streamline the setup by using thresholds of 0.6 and 0.7, reducing the
number of bootstrap samples to 30.

To summarize the results, we identify the most frequently occurring edge
endpoints across different experimental setups.\footnote{The detailed
  proportion of each edge endpoint occurrence is shown in
  Section~\ref{sec-propmatrix}.} This ensures that only stable and
consistent edges are retained, providing a clearer and more reliable
understanding of the relationships between precarity factors and
depression. Refer to Figure~\ref{fig-workflow} for the analysis
workflow.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{img/simsetup.png}

@@

>>>>

\date{2024-12-30}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

@@

After preprocessing, the HELIUS dataset comprises 21,628 samples. Along
with the five precariousness factors, we also compute an overall
precarity score as the combined value of these factors. Additionally,
PHQ-9 scores are included to represent depression, both as a total sum
score and as individual symptom scores. In the subsequent causal
discovery analysis, we examine the relationship between depression and
precarity using both aggregated sum scores and their individual-level
representations. Refer to Figure~\ref{fig-dist} for the overall
distributions of the variables used in the analysis.

@@

\caption{\label{fig-dist}Distributions of variables with density
overlay. \emph{P.emp} = employment precariousness; \emph{P.hou} =
housing precariousness; \emph{P.soc} = social precariousness;
\emph{S.fin} = recent financial stressors; \emph{S.rel} = recent
relational stressors; \emph{Precarity} = overall precarity;
\emph{PHQsum} = PHQ-9 sum score; \emph{anh} = anhedonia; \emph{app} =
appetite; \emph{con} = concentration; \emph{dep} = depressed mood;
\emph{ene} = energy; \emph{glt} = guilty; \emph{mot} = motor; \emph{sui}
= suicidal}

@@

We analyze the causal structure using three approaches: examining the
relationship (1) between the five individual precarity factors and the
PHQ sum score, which represents overall depression severity; (2) between
individual symptoms and precarity factors; and (3) between individual
symptoms and overall precarity (the sum of the five precarity factors).
Using the PHQ sum score simplifies the analysis by reducing
dimensionality, offering computational efficiency and a broad,
interpretable perspective on the relationship between depression and
precarity factors. In contrast, including individual symptoms provide a
more detailed understanding, capturing the diverse ways symptoms respond
to different precarity factors. However, this approach introduces
methodological challenges, such as non-standard distributions and the
complexities of increased dimensionality. Finally, examining individual
symptoms in relation to the overall precarity score complements the
analysis by revealing how symptoms collectively connect to general
precarity. By integrating these three approaches, we balance simplicity
and granularity, ensuring robust findings.

To address sensitivity to parameter values and enhance the robustness of
our results, we implement multiple settings combined with bootstrapping.
For each of the 100 bootstrap samples, we estimate causal graphs and
retain only the edges and directions that exceed predefined thresholds.
The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds
× 2 CI tests), applied across three algorithms. Each combination is
repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per
algorithm. For analyses involving individual symptom variables, we
simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
the number of bootstrap samples to 30 to ease computational demands.

To summarize the results, we identify the most frequently occurring edge
endpoints across different experimental setups.\footnote{The detailed
  proportion of each edge endpoint occurrence is shown in
  Section~\ref{sec-propmatrix}.} This ensures that only stable and
consistent edges are retained, providing a clearer and more reliable
understanding of the relationships between precarity factors and
depression. See Figure~\ref{fig-workflow} for an overview of the
analysis workflow.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{img/simsetup2.png}

@@

Lastly, we examine the relationships between individual symptoms and
overall precarity, represented as the sum of five precariousness
factors. This approach completes the analysis by providing a perspective
on how individual symptoms relate to overall precarity. It complements
the previous analysis, which focused on individual symptoms and
individual precarity factors, by potentially capturing distributed
relationships across different precarity factors through their
aggregation into a single overall score. As before, black solid lines
indicate edges agreed upon by both FCI and CCI, while gray dashed lines
represent edges where the two algorithms disagree. Circles at the
endpoints of gray dashed edges denote unresolved directionality.
Additionally, navy dashed edges highlight relationships that appear
exclusively in graphs generated using Gaussian CI tests.

Compared to Figure~\ref{fig-sym}, this analysis reveals greater
disagreement between FCI and CCI regarding edge directionality, as
evidenced by more dashed edges. FCI tends to estimate more undirected
edges (circles), while CCI more frequently assigns directed arrows
between symptom-symptom relationships. For a detailed breakdown, see
Section~\ref{sec-propmatrix}, which shows the proportion of symbols
produced by each algorithm. When examining overall precarity, we observe
that more symptoms are now connected to the precarity score
(\emph{Precarity}) compared to the individual precarity factor analysis
in Figure~\ref{fig-sym}. In addition to \emph{glt}, \emph{slp},
\emph{anh}, and \emph{sui}, which are previously shown to be connected
to individual precarity factors, \emph{mot}, \emph{app}, and \emph{dep}
are now identified as having causal relationships with overall
precarity. Furthermore, when using the Gaussian CI test, con is also
shown to be causally related to overall precarity. Notably, all these
connections are directed towards precarity (o→), indicating that overall
precarity is not the cause of symptoms. The exception is the
bidirectional edge between \emph{sui} \textless-\textgreater{}
\emph{Precarity}, which suggests a connection via latent confounders.
FCI predominantly estimates directed arrows (→) for symptom-precarity
relationships, suggesting that symptoms are driving precarity. With
fewer variables (reduced from 14 to 9), this approach appears to have
slightly more statistical power, enabling algorithms to estimate
relationships with greater confidence compared to the previous analysis
involving individual symptoms and individual precarity factors.
